\thesistitle

### 2025-07-24 manuscript/main.tex
multicol abbr.  fix foramtting
Conclusion (Requirement fullfillment)

\noindent
Ich erkläre weiters, dass ich mich generativer KI-Tools lediglich als Hilfsmittel bedient habe und in der vorliegenden Arbeit mein gestalterischer Einfluss überwiegt. In der "Anerkennung"(Acknowledgements) habe ich alle generativen KI-Tools gelistet, die verwendet wurden, und angegeben, wo und wie sie verwendet wurden. Für Textpassagen, die ohne substantielle Änderungen übernommen wurden, haben ich jeweils die von mir formulierten Eingaben (Prompts) und die verwendete IT-Anwendung mit ihrem Produktnamen und Versionsnummer/Datum angegeben.

@@

\pagenumbering{roman}
\cleardoublepage
% — Right-aligned, large title —
\phantomsection
\begin{flushright}
  {\Huge\bfseries Abstract\par}
\end{flushright}
\addcontentsline{toc}{chapter}{Abstract}
% Push the block down
\vspace*{3cm}

@@

\clearpage
\chapter{Acknowledgements}

\noindent I would like to thank my supervisory team - Michele Chiari, Francesco Pontiggia and Ezio Bartocci - for generously sharing their expertise throughout this project.
Their patient guidance on implementation details, contextual foundations, and domain-specific nuances of the application was invaluable. Whenever I encountered a technical
hurdle or needed deeper insight into the domain, they were always ready with thoughtful advice and resources.
\vspace{1ex}
\noindent I would also like to acknowledge the use of ChatGPT 4o-mini-high during the preparation of this thesis. I used it to enhance my writing by correcting grammar
and spelling errors, as well as improve the overall tone and wording.
Furthermore, the AI-tools was used as a consultation on common topics and for creating the Biblatex citations.
All final texts and substantive content, however, reflect my own efforts and have been critically reviewed by me.
\clearpage

\chapter{Abbreviations}
\begin{description}
  \item[API] Application Programming Interface
  \item[AST] Abstract Syntax Tree
  \item[BNF] Backus-Naur Form
  \item[BOOL] Boolean
  \item[CAV] Computer-Aided Verification
  \item[CFG] Context-Free Grammar
  \item[CLI] Command-Line Interface
  \item[CPS] Cyber-Physical System
  \item[CS] Computer Science
  \item[CST] Concrete Syntax Tree
  \item[DFA] Deterministic Finite Automaton
  \item[DI] Dependency Injection
  \item[DSL] Domain-Specific Language
  \item[EBNF] Extended Backus-Naur Form
  \item[ESM] ECMAScript Module
  \item[JSON-RPC] JavaScript Object Notation -- Remote Procedure Call
  \item[LL] Left-to-right, Leftmost derivation
  \item[LR] Left-to-right, Rightmost derivation
  \item[LSP] Language Server Protocol
  \item[LTL] Linear Temporal Logic
  \item[MCMC] Markov Chain Monte Carlo
  \item[OPA] Operator Precedence Automaton
  \item[OPM] Operator Precedence Matrix
  \item[PDA] Pushdown Automaton
  \item[PEG] Parsing Expression Grammar
  \item[POTL] Procedence Operator Temporal Logic
  \item[PPL] Probabilistic Programming Language
  \item[SMC] Statistical Model Checking
  \item[SMT] Satisfiability Modulo Theories
  \item[URI] Uniform Resource Identifier
  \item[VI] Variational Inference
  \item[VS] Visual Studio
  \item[VSCE] Visual Studio Code Extension
  \item[WS] Web Services
  \item[XML] eXtensible Markup Language
\end{description}

@@

\chapter{section}{Expanded Code Listings}
\inputminted[linenos, breaklines]{yaml}{samples/mini-prob.langium}
\captionof{listing}{Langium grammar definition of MiniProb (LL(2) parsable)}
\label{list:langium-grammar}

>>>>

\noindent
Ich erkläre weiters, dass ich mich generativer KI-Tools lediglich als Hilfsmittel bedient habe und in der vorliegenden Arbeit mein gestalterischer Einfluss überwiegt. In der "Anerkennung"(Acknowledgements \ref{sec:ack}) habe ich alle generativen KI-Tools gelistet, die verwendet wurden, und angegeben, wo und wie sie verwendet wurden. Für Textpassagen, die ohne substantielle Änderungen übernommen wurden, haben ich jeweils die von mir formulierten Eingaben (Prompts) und die verwendete IT-Anwendung mit ihrem Produktnamen und Versionsnummer/Datum angegeben.

@@

\clearpage
\thispagestyle{empty}
\begin{flushright}
  {\LARGE\bfseries Acknowledgements\par}
\end{flushright}
\vspace{5cm}

\noindent I would like to thank my supervisory team - Michele Chiari, Francesco Pontiggia and Ezio Bartocci - for generously sharing their expertise throughout this project.
Their patient guidance on implementation details, contextual foundations, and domain-specific nuances of the application was invaluable. Whenever I encountered a technical
hurdle or needed deeper insight into the domain, they were always ready with thoughtful advice and resources.
\vspace{1ex}

\noindent I would also like to acknowledge the use of ChatGPT 4o-mini-high during the preparation of this thesis. I used it to enhance my writing by correcting grammar
and spelling errors, as well as improve the overall tone and wording.
Furthermore, the AI-tool was used as a consultation on common topics and for creating the Biblatex citations.
All final texts and substantive content, however, reflect my own efforts and have been critically reviewed by me.

\pagenumbering{roman}

\cleardoublepage
% — Right-aligned, large title —
\phantomsection
\begin{flushright}
  {\Huge\bfseries Abstract\par}
\end{flushright}
\addcontentsline{toc}{chapter}{Abstract}
% Push the block down
\vspace*{3cm}

@@

\chapter{Abbreviations}
\begin{multicols}{2}
  \begin{description}
  \item[API] Application Programming Interface
  \item[AST] Abstract Syntax Tree
  \item[BNF] Backus-Naur Form
  \item[BOOL] Boolean
  \item[CAV] Computer-Aided Verification
  \item[CFG] Context-Free Grammar
  \item[CLI] Command-Line Interface
  \item[CPS] Cyber-Physical System
  \item[CS] Computer Science
  \item[CST] Concrete Syntax Tree
  \item[DFA] Deterministic Finite Automaton
  \item[DI] Dependency Injection
  \item[DSL] Domain-Specific Language
  \item[EBNF] Extended Backus-Naur Form
  \item[ESM] ECMAScript Module
  \item[JSON-RPC] JavaScript Object Notation -- Remote Procedure Call
  \item[LL] Left-to-right, Leftmost derivation
  \item[LR] Left-to-right, Rightmost derivation
  \item[LSP] Language Server Protocol
  \item[LTL] Linear Temporal Logic
  \item[MCMC] Markov Chain Monte Carlo
  \item[OPA] Operator Precedence Automaton
  \item[OPM] Operator Precedence Matrix
  \item[PDA] Pushdown Automaton
  \item[PEG] Parsing Expression Grammar
  \item[POTL] Procedence Operator Temporal Logic
  \item[PPL] Probabilistic Programming Language
  \item[SMC] Statistical Model Checking
  \item[SMT] Satisfiability Modulo Theories
  \item[URI] Uniform Resource Identifier
  \item[VI] Variational Inference
  \item[VS] Visual Studio
  \item[VSCE] Visual Studio Code Extension
  \item[WS] Web Services
  \item[XML] eXtensible Markup Language
\end{description}
\end{multicols}

@@

\chapter{Expanded Code Listings}
\inputminted[linenos, breaklines]{yaml}{samples/mini-prob.langium}
\captionof{listing}{Langium grammar definition of MiniProb (LL(2) parsable)}
\label{list:langium-grammar}

### 2025-07-24 manuscript/part2.tex
Conclusion (Requirement fullfillment)

reqjuiremnts fullfilled, code completeion lacks in basic stucutrues only for linked references.
conlsuio: mention no dode comülpeteion for structures only refs, not optimal and how it would be implemetned

>>>>

In this work, a Visual Studio Code extension for MiniProb was developed that, by and large, satisfies the functional requirements set out at the outset.
Syntax highlighting for keywords, operators, literals and identifiers is fully realized; data type validation and real-time semantic analysis - including detection of
undefined variables and invalid constructs - are wholly supported by a combination of custom Langium services and its built-in validation pipeline. Error messages
appear immediately as users edit their code, albeit with the caveat that some default Chevrotain diagnostics remain terse when rules are incomplete. Basic code
completion is operational for referenced names and variables, but full support for language constructs awaits the implementation of a dedicated CompletionProvider.

On the non-functional side, the extension responds to edits within 5-200ms for average sized files, comfortably within the 100-250ms target, and even large files
(approx. 10000 lines) load and validate within one second - though this does not hold on weaker machines (>=2 seconds). Compatibility with the latest VS Code
release is ensured via up-to-date Langium and LSP dependencies. Comprehensive installation and usage instructions are provided both in this thesis and in the project's
GitHub repository, facilitating future maintenance and feature extension. Overall, the extension meets its core objectives, delivering a responsive, semantically aware
authoring environment for probabilistic DSLs; remaining work centers on enriching code-completion capabilities to close the final gap in our functional specification.
